---
title: "Data Wrangling in R"
author: "Boston University"
date: "January 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Science Workflow

1. Import
2. Clean (tidy)
3. Understand:
 + Transform
 + Visualize
 + Model
4. Communicate


The tidyverse is an opinionated collection of R packages designed for data science. 
All packages share an underlying design philosophy, grammar, and data structures. 


### Core tidyverse packages 
* **readr** - import data
* **tidyr** - for data "tidying"
* **dplyr** - data manipulation
* **stringr** - string manipulation
* **forcats** -factors handling
* **ggplot2** - data visualization
* **tibble** - modern representation of a data frame
* **purrr** - functional programming

### *** Other related packages ***
* **hms** - time manipulation
* **lubridate** - dates/times manimpulation
* **DBI** - databases
* **haven** - data imported from SPSS, SAS and Stata 
* **readxl** - excel data import
* **rvest** - web scaping

Tidyverse packages can all be installed together:
```{r eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

We will be using two packages - dplyr and readxl. If these packages have not been installed on your computer, execute:
```{r eval=FALSE}
install.packages("readxl")
install.packages("dplyr")
```
<br><br>
First we need to load packages into R environment
```{r load], message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
```

## Importing dataset from Excel worksheet
Reading data saved in Excel format often comes with some challenges:
```{r }
visit <- read_excel("MedData.xlsx", sheet = "Visits")
visit
```
<br><br>
We can see that the file contains some extra records at the top and at the bottom of the sheet.
Use options to read_excel() function to skip these extra lines:
```{r }
visit <- read_excel("MedData.xlsx", 
                    sheet = "Visits", 
                    skip = 3,  # how many lines to skip at the top of the sheet
                    n_max = 13)  # how many observations to read
visit
```

Let's take a look at the type of each record to make sure that each column was read correctly
This can be done with `str()` function.
However dplyr package has also handy glimpse() function that has a nicely formatted output
```{r }
str(visit)
glimpse(visit)
```


Notice that DBP column is marked as "character" column even though it contains numeric values.
This normally means that there are some missing values that were coded in non-standard way, so 
R did not recognize them as missing values.
In this case you should examine the column and try to identify the missing values:
```{r }
min(visit$DBP)
```
<br><br>
Now we can go back and read file again adding `##N/A` as a Missing Data code:
```{r }
visit <- read_excel("MedData.xlsx", 
                    sheet = "Visits", 
                    skip = 3,
                    n_max = 13,
                    na=c("","NA","##N/A"))
glimpse(visit)
```

## Pipe symbol `%>%`

"Pipe" symbol in R was introduced in magrittr package and then it was also implemented in dplyr package - package we will use today
It allows to send the output of one function as an input object to another function:
```{r }
visit %>% glimpse()
visit %>% summary()
```
<br><br>
The following chain of functions:
```{r }
sort(unique(visit$`Patient ID`), decreasing=TRUE)
```
can be rewritten as
```{r }
visit$`Patient ID` %>% 
  unique()  %>%            # find unique values
  sort(decreasing=TRUE)    # sort  in descending order
```


## Cleaning and analysing a dataset
 A dataset is a collection of values, that are usually either:
 
* numbers (if quantitative),  
* strings (if qualitative),
* logical (if binary).

Every value belongs to a variable and an observation. 

A variable contains all values that measure the same underlying attribute (like phone, age) across units. 

An observation contains all values measured on the same unit (like a person, or a day, or an event) across attributes

#### Clean or Tidy Data:
* Each variable forms a column
* Each observation forms a row
* Each type of observational unit forms a table

## Examples of messy datasets
* Column headers are values, not variable names
* Multiple variables are stored in one column
* Variables stored in both rows and columns
* Multiple types of observational units are stored in the same table
* A single observational unit is stored in multiple tables.

### dplyr package workflow

`dplyr` R package allows us to:

* Rename variables ( `rename()` )
* Filter observations by their values ( `filter()` )
* Reorder rows ( `arrange()` )
* Select specific columns/variables ( `select()` )
* Create new variables with functions of existing variables ( `mutate()` )
* Summarise ( `summarise()` )


### Renaming columns

# We can select columns in a Data Frame using "$" symbol, i.e:

```{r }
visit$DBP
```
<br><br>
# It is not as easy to do so when the column name contains spaces or other special characters:
```{r }
visit$`Patient ID`
```
<br><br>
In some cases in makes sense to rename some columns to make it easy to work :
```{r }
visit.clean <- visit %>%
  rename(id="Patient ID", 
         admission="Admission date",
         discharge="Discharge date",
         pulse="Heart Rate")
names(visit.clean)
```


### Working with strings

```{r }
# Converting strings to upper and lower case
toupper(visit.clean$Allergies)
# or using pipe operator
visit.clean$Allergies %>% toupper()

#Search character vector for a specific pattern:
grepl ("pain", visit.clean$Symptoms,  ignore.case=TRUE)

# Find all strings that start with "chest"
grepl ("^chest", visit.clean$Symptoms,  ignore.case=TRUE)

# Find all strings that end with "pain"
grepl ("^chest", visit.clean$Symptoms,  ignore.case=TRUE)

# Find all strings that contain either fever or pain (or both)
grepl ("fever|pain", visit.clean$Symptoms,  ignore.case=TRUE)
```
<br><br>
`grep( value=FALSE)`: returns a vector of indecies of element where pattern is found<br>
`grep( value=TRUE)`: returns a vector of indecies of element where pattern is found
```{r }
grep ("fever", visit.clean$Symptoms,  ignore.case=TRUE, value=FALSE)
grep ("fever", visit.clean$Symptoms,  ignore.case=TRUE, value=TRUE)
```
<br><br>
`sub()` function can be used to substitute the first occurance of a pattern with another string
```{r }
sub (",", ";", visit.clean$Symptoms)
```

<br><br>
`gsub()` function can be used to substitute all occurances of a pattern with another string
```{r }
gsub (",", ";", visit.clean$Symptoms)
```

### Working with dates


```{r }
class(visit.clean$admission)

curr.time <- Sys.time()  # get current date and time
curr.date <- Sys.Date()  # get current date 
str(curr.time)           # view the structure of an object
class(curr.time)         # view type of an object

# some systems do not have timezone set up
Sys.timezone()

# convert character string to POSIXlt
t1=as.POSIXct("2019-01-29 11:30:00", "%Y-%m-%d %H:%M:%S", tz="EST")
str(t1)  

# list of available Time Zones
#OlsonNames()  
```         


There are a number of handy packages that have various functions to work with dates:
* lubridate
* chron

Some more examples working with dates and times can be found in the following R script
[rcs.bu.edu/examples/r/timesSeries/dateTime.R](http://rcs.bu.edu/examples/r/timesSeries/dateTime.R)
<br><br>
Here we will calculate the length of stay of each patient in the hospital:
```{r }
visit.clean$discharge - visit.clean$admission
#The result variable has "difftime" class:
stay <- visit.clean$discharge - visit.clean$admission
class(stay)
#To convert it to a numeric value use as.numeric() function:
as.numeric(visit.clean$discharge - visit.clean$admission)
```


### Filtering rows
```{r }
visit.clean %>% filter( !is.na(DBP) )
```

R logical operators:

* `>  >=`
* `==  !=`
* `<   <=`
* `is.na()   !is.na()`
*  `%in%`


#### Examples
```{r }
# Select only those records for which pulse columns have values 100 and greater
visit.clean %>% filter( pulse > 100 )

# Select only those records for which DBP is less than 60 or SBP is greater than 120
visit.clean %>% filter( DBP < 60 | SBP > 120)

# Select only those records for which Temperature is greater than 99 and Symptoms include "fever"
visit.clean %>% filter( Temperature > 99 &  grepl("fever", Symptoms, ignore.case=T)   )

```


### Selecting Columns

```{r }
visit.clean %>% select(id, Temperature:pulse)
```
<br><br>
dplyr comes with a set of helper functions that can help you select groups of variables inside a select() call:

* `starts_with("XY")`: every name that starts with "XY",
* `ends_with("XY")`: every name that ends with "XY",
* `contains("XY")`: every name that contains "XY",
* `matches("XY")`: every name that matches "XY", where "XY" can be a regular expression,
* `num_range("x", 1:15)`: the variables named x01, x02, x03,..., x15,
* `one_of(XY)`: every name that appears in x, where XY is a character vector.

```{r }
visit.clean %>% select( id, ends_with("BP") )
```


### Modifying existing and/or creating a new variables

```{r }
visit.clean %>% mutate( Temperature = (Temperature-32)*5/9, stay = as.numeric(discharge - admission) )
```


#### Examples

```{r }
# Create a new column MAP which is equal to SBP/3 + 2*DBP/3
visit.clean %>% mutate( MAP=SBP/3 + 2*DBP/3 ) 
```
<br><br>
Let's put it all together:<br>
Use visit.clean dataframe as input and

* - select only those columns where Temperature is greater than 99F
* - select columns ID, DBP and SBP
* - calculate new variable MAP
```{r }
res <- visit.clean %>%
            filter( Temperature > 99 ) %>%
            select( id, DBP, SBP ) %>%
            mutate( MAP = SBP/3 + 2*DBP/3 ) 
 res



###  Calculating summaries

 ```{r }
visit.clean %>% summarise( N = n(), max.t = max(Temperature), min.T = min(Temperature) )
```
<br><br>
**bilt-in functions often used within summarise()**:

* averages: `mean(), median()`
* spread: `sd(), IQR(), mad()`
* range: `min(), max()`
* count: `n(), n_distinct()`


#### Examples:

```{r }
# Calculate the number of distinct patients
visit.clean %>% summarise( N = n_distinct(id) ) 
```


###  Group by one or more variables

```{r }
visit.clean %>% group_by(id) %>% summarise( ave.pulse = mean(pulse) )
```
<br><br>
Useful functions often used with group_by():

* `first()` 
* `last()`
* `nth()`

```{r }
# For each patient select the first record and find the lenght of stay
visit.clean %>% group_by(id) %>% summarise( first(discharge - admission) )
```


###  Sorting dataframe by one or more variables
```{r }
visit.clean %>% arrange(id, admission)
```


##  Joining 2 dataframes


First let's read the data for each patient:
```{r }
pinfo <- read_excel("MedData.xlsx", sheet = "Patient Data")

#let's make sure the date is well formatted
pinfo %>%head()
pinfo %>%glimpse()
pinfo %>%summary()
```

Now we want to have a single dataframe that contains patient infomration and patient visit informtaion

There are a number of `join` functions in `dplyr` package:

* `inner_join()`
* `left_join()`
* `right_join()`
* `full_join()`
* `semi_join()`  - return all rows from x where there are matching values in y, keeping just columns from x.
* `anti_join()`  - return all rows from x where there are not matching values in y, keeping just columns from x.

```{r }
#Let's try full join of both dataframes we have
full_join(visit.clean, pinfo, by = c("id"="ID"))

# left join
result <- visit.clean  %>% left_join(pinfo, by = c("id"="ID"))
result



# Calculate the mean stay of the hospital for Male and Female patients:
result %>%
  filter( !is.na(Gender) ) %>%            
  mutate( stay = as.numeric(discharge-admission) ) %>%     
  group_by ( Gender ) %>%         
  summarize( ave.stay = mean(stay) )  
  
```

## Converting Wide dataframes to long dataframes and back

```{r }
library(tidyr)
city.temps <- data.frame( time = as.Date('2018-09-03') + 0:4,
                          Boston = rnorm( 5, 75, 3),
                          Denver = rnorm( 5, 85, 5),
                          SanFrancisco = rnorm(5, 80, 5),
                          Austin = rnorm( 5, 90, 5) )
city.temps

# Use gather to rearrange a table into a tidy data.frame:
# gather(data, key = "key", value = "value", ..., na.rm = FALSE,
# convert = FALSE, factor_key = FALSE)
city.temps2 <- gather( city.temps, 
                       key = "City",
                       value = "Temperature",
                       -time,  # collection of the columns,
                       factor_key = TRUE) # if key variable needs to be converted to a factor
# or
city.temps2 <- gather( city.temps, 
                       key = "City",
                       value = "Temperature",
                       Boston: Austin,  # collection of the columns,
                       factor_key = TRUE) # if key variable needs to be converted to a factor
#or
city.temps2 <- gather( city.temps, 
                       key = "City",
                       value = "Temperature",
                       c("Boston","Denver","SanFrancisco","Austin"),  # collection of the columns,
                       factor_key = TRUE) # if key variable needs to be converted to a factor


city.temps2
glimpse(city.temps2)
```
<br><br>
Sometimes it is useful to be able to
perform the opposite operation: convert long format dataframe into wide representation

```{r }
city.temps3 <- spread( city.temps2, City, Temperature)
city.temps3
```


