---
title: "R1-Intro"
author: "Research Computing Services"
date: "Boston University"
output: html_document
---

<br><br>

### R as a scientific calculator

```{r }
2 + 3    # addition
```


```{r }
2 ^ 3    # power
log(2)   # built-in functions
```

<br><br>

### Variable (objects) definition
The traditional R way to assign a value to a variable is to use symbol <em><-</em>, but <em>=</em> can be used too:
```{r }
a <- 3
b = 2 ^ (2 + 2) 
```
To view the value of a variable use command *print*:
```{r }
print(b)
```
You can also simply print the name of the R object at the prompt of the command line:
```{r }
b
```


R is case sensitive:
```{r }
A <- 7
print(A-a) 
```
Avoid using names c, t, cat, F, T, D and other built-in functions and constants

The variable name can contain letters, digits, underscores and dots and start with the letter or dot. The variable name cannot contain dollar signs or other special characters as they will be used to built more complex R structures like lists and data frames.
Make your variable names descriptive - use period (<em>.</em>) or underscore (<em>_</em>) to separate words.

There are numeric, character and logical variable types in R:
```{r }
element <- "Au"   # character variable (single or double quotes are OK)
number <- 79      # numeric variable
metal <- TRUE     # logical variable
```

### Vectors
Vector is an list of R objects of the same type:
```{r }
seq<- c ("CCTTCCTAC", "GAAATAGT", "ACTGGGGGATACG")  # DNA sequences
len <- c(9, 8, 13)
```

Vectors can be defined in a veriety of ways. Above we used function **c()** to combine a few values into a vector. There are many other methods to create a vector in R. Here are a few examples. The outer brackets around the expression forces the result of the operation to be printed:
```{r }
# define a range of values: 
( ids <- 101:150 )
# sequence definition
( steps <- seq(from=0, to=5, by=0.1) )
# repeat same value
( score <- rep(1, 10) )
# 25 normally distributed values
( random <- rnorm(25) )
```

The last 3 vectors were defined using R functions.

### Getting Help in R

To access help file for the R function with a known name, type question mark with a function name:
```{r, eval=FALSE}
?seq
#or
help(seq)
```

If the name function is not know, use **??** to search R help articles:
```{r, eval=FALSE}
??"standard deviation"
#or
help.search("standard deviation")
```

### Vector arithmetic:
 R vectors can be added together, subtracted, multiplied, devided etc.:
```{r }
# systolic blood pressure values
SBP <- c(96, 110, 100, 125, 90 )

# diastolic blood pressure
DBP <- c(55, 70, 68, 100, 50)

# calculate MAP (mean arterial pressure)
MAP <- SBP / 3 + 2 * DBP / 3
print(MAP)
```


### Vector Slicing
```{r }
temp <- c(36.6, 38.2, 36.4, 37.9, 41.0, 39.9, 36.8, 37.5)    # define a numeric vector
temp[2]         # returns second element 
temp[2:4]       # returns second through 4th elements inclusive
temp[c(1,3,5)]  # returns 1st, 3rd and 5th elements
temp[-2]        # returns all but 2nd element
temp[c(TRUE, FALSE, TRUE, FALSE, FALSE,FALSE, TRUE, FALSE)]   # returns 1st, 3rd, and 7th elements

#compare each element of the vector with a value
temp < 37.0

#return only those elements of the vector that satisfy a specific condition
temp[ temp < 37.0 ]    
```

### Basic functions to analyze vector

```{r}

which.max(temp)  # find the (first)maximum element and return its index
which.min(temp)
which(temp >= 37.0) # find the location of all the elements that satisfy a specific condition
```

There are a number of other useful functions:

**max(x),   min(x),     sum(x)** <br>
**mean(x),  sd(), median(x),  range(x)** <br>
**var(x)**       - simple variance <br>
**cor(x,y)**     - correlation between 2 vectors  <br>
**sort(x), rank(x), order(x)**  <br>
**cumsum(), cumprod(x), cummin(x), cumprod(x)**  <br>
**duplicated(x), unique(x)**  <br>
**summary()**



### Reading Data
```{r}
df <- read.csv( "http://scv.bu.edu/examples/r/tutorials/Salaries.csv")
```
All the data from the file are stored in the R object *df* called **data frame**.

The following webpage is CRAN official documentation about reading inputs of various formats:<br>
[https://cran.r-project.org/doc/manuals/r-release/R-data.html](https://cran.r-project.org/doc/manuals/r-release/R-data.html)


### Explore the data #

```{r}
#numeric data exploratory analysis
min(df$salary)
max(df$salary)
range(df$yrs.service)
summary(df$salary)

#view the data
hist(df$salary)

#another way to look at the data
boxplot( df$salary )

#view qq-plot to see if the data normaly distributed
qqnorm(df$salary); qqline(df$salary)

#Shapiro-Wilks test of normality
shapiro.test(df$salary)
#Null-hypothesis - the sample comes from a normal distribution
#Alternative hypothesis - the sample does not come from a normal distribution
#p-value < 0.05 - reject the Null-hypothesis - data are not normally distributed


#categorical data analysis
summary(df$rank)
summary(df$sex)
```

### Dataframe slicing

```{r}
# dataframe slicing (accessing elements)
df[3,5]    # element, from the 3rd row, 5th column
df[3,]     # all elements from the 3rd row
df[ ,5]    # all elements in the 5th column 

df$sex  # accessing elemnts using variable name
df[df$sex == "Male", ]  # list all the rows in the dataset for which variable ed is equal to specific value

#Create a new dataframe as a subset of the original one
disc.B <- df[df$discipline == "B", ] 
head(disc.B)
```

###   data analysis         


We would like to complare to groups of the dataframe. Let's compare the salaries of 
* Assistant Professor, 
* Associate Professor,
* Professor

```{r}
boxplot(salary ~ rank, data=df)

#get mean value for each subgroup
mean( df$salary[ df$rank == "AsstProf" ] )
mean( df$salary[ df$rank == "AssocProf" ] )
mean( df$salary[ df$rank == "Prof" ] )

#There is an easier way to perform the above calculation for each value of a categorical variable:
tapply( df$salary,  df$rank, mean)
```


Is this difference is actually significant to state that women's salary is lower than men's
We will perform the analysis of variance (anova test)
```{r}
aov.res <- aov( salary ~ sex, data=df )
summary(aov.res)
```

Since p-value is < 0.05 we reject the Null hypothesis of equal means
F - is the ration of two mean square values. If the Null hypothesis is true, 
we expect F to be close to 1.  large F ratio means that variation among group means
is larger than we expect to see by chance

<br><br>
The difference in salary for various disciplines in this dataset is even more pronounced:
```{r}
boxplot( salary ~ sex + discipline, data=df  )
aov.res <- aov( salary ~ sex + discipline, data=df )
summary(aov.res)
```


Multiple Comparisions
```{r}
boxplot( salary ~ rank, data=df  )
plot.design( salary ~ rank, data=df )
aov.res <- aov( salary ~ rank, data=df )
summary(aov.res)
```
Here we definetly reject the Null hypothesis, but we cannot state that all groups have different means

To compare means between each subgroup - perform Tukey honestly significant difference test
```{r}
TukeyHSD(aov.res)
```
There is a significant difference between Professor and the other 2 groups, but
we cannot reject a Null hypothesis comparing the mean salaries of Assistant Professor and Associate Professor


### Analysis of 2 numeric variables
```{r}
plot(df$yrs.service, df$yrs.since.phd)

# Linear regression: fit linear model
lm.fit <- lm (yrs.since.phd ~ yrs.service, data = df) 
summary(lm.fit)   
coef(lm.fit)   # access coefficients
resid(lm.fit)  # access residual errors
fitted(lm.fit) # predicted values for dt$ap
abline (lm.fit)

#predict
predict (lm.fit, data.frame(yrs.service=10:20) )
```

### Missing values in R

```{r}
x <- c(734, 145, NA, 456, NA)    # define a numeric vector
is.na(x)              # check if the element in the vector is missing
which(is.na(x))       # which elements are missing
anyNA(x)              # are there any missing data
```

x == NA   does not work! - missing value cannot be compared to anything

```{r}
# By default mamy statistical functions will not compute if the data contain missing values
mean(x)


# Read help topic for the function
#?mean

#Perform computation removing the missing data
mean(x, na.rm=TRUE)
```





